DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-04-08 manuscript/manuscript.Rmd
cross validation XY figure

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs. \ref{fig:cv1} to \ref{fig:cv10}).
<!--I'm confused about why you are talking about branch lengths here... did you use branch lengths? Yes, on the PATHd8 cross validation, but you are right, it is confusing, so I removed that first sentence-->

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
We also demonstrate the usage of PATHd8 [@Britton2007] as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using `datelife`'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
<!-- What made it unsuccessful in the other cases? -->
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
<!-- ??? Node ages or maximum ages?? -->
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

>>>>

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
To perform a cross validation analysis of the DateLife workflow, we took tree topologies from source chronograms and used age data from all other source chronograms as fixed calibrations with the program BLADJ (Fig. \ref{fig:cvXY}).
<!--I'm confused about why you are talking about branch lengths here... did you use branch lengths? Yes, on the PATHd8 cross validation, but you are right, it is confusing, so I removed that first sentence-->

*I still need to double check results to confirm or change any of the following descriptions:*

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvXY}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvXY}).
<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
We also demonstrate the usage of PATHd8 [@Britton2007] as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using `datelife`'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
<!-- What made it unsuccessful in the other cases? -->
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
<!-- ??? Node ages or maximum ages?? -->
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

### 2022-04-08 manuscript/manuscript.tex
cross validation XY figure

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4379}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs. \ref{fig:cv1} to \ref{fig:cv10}).

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using \texttt{datelife}'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
\caption{Cross validation of first source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
\caption{Cross validation of second source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv2}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
\caption{Cross validation of third source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv3}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
\caption{Cross validation of fourth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv4}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
\caption{Cross validation of sixth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv6}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
\caption{Cross validation of seventh source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv7}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
\caption{Cross validation of eight source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv8}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
\caption{Cross validation of ninth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv9}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
\caption{Cross validation of tenth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv10}
\end{figure}

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4391}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

To perform a cross validation analysis of the DateLife workflow, we took tree topologies from source chronograms and used age data from all other source chronograms as fixed calibrations with the program BLADJ (Fig. \ref{fig:cvXY}).

\emph{I still need to double check results to confirm or change any of the following descriptions:}

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvXY}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvXY}).
We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using \texttt{datelife}'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
\caption{Results from cross validation analysis.}
\label{fig:cvXY}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
